Collaborative Visual Analysis on the Web with RCloud

### 2015-06-26 interviews.tex
topic sentences; trim boring stuff
in "evaluation" section
also remove all comments to make this easier to edit

\section{Interview Study\label{sec:interviews}}

To evaluate the effectiveness of RCloud, we interviewed 13 current or
recent users of RCloud. 9 are data analysts, and 4 build tools for data
analysts and business needs.
% no paragraphbreak needed! scn
The subjects differed in their evaluation of its benefits and disadvantages,
as described below.

%Analysts were more annoyed by the cell interface and its perceived
%slowness than were web/tool developers. Analysts
%switching to R from another language were more likely to use RCloud for
%exploratory data analysis. Web/tool developers were more likely to see the
%need for parameterization.

@@

%% Wendy's usage is typical:
%% \begin{quote}
%% I've been sharing my notebooks in order for other people to see what
%% I've done. It's very convenient for that purpose. [...] We haven't
%% been editing them together, but we've been just sharing, and I've been looking
%% at things that other people do. So there's no collaborative aspect of it, like
%% developing the code, but there's sharing aspects for other people to see and for
%% me to see other people's notebooks.
%% \end{quote}

@@

Eric develops packages for analysts, and uses RCloud ``more for sharing code with
other people, and for doing tutorials for {\em iotools} or {\em hmr}'', his packages.
``I want people to see how the package works, so I clearly want them
to see the code... I write it just like I would write GitHub
Markdown, where you have little code snippets and text, but RCloud lets me
actually run the snippets [and display the results].'' He also uses RCloud for
describing and debugging data sources.

Some users, who tried RCloud and were not able to continue for organizational
reasons, miss certain capabilities. Leith ``[likes] \comment{the concept of} being
able to create notebooks and share them... A wiki is not the best way for
communicating results - it's \comment{kind of} like writing a blog post with very limited
functionality. I have to save every picture and post it as an image.'' She said,
``I can't share all of the code because it would just get crowded and wouldn't
look right on a wiki.'' Iris explains, ``If I could make a folder on RCloud and
have Python notebooks and also Pig notebooks there, and execute them from RCloud,
that would be much better than my current [environment], because that would free
me from manual documentation and version control and also telling people where
my code was. It would be just, hey, go look on RCloud, here's the stuff.''

%% Kaylee: ``being able to share notebooks or codes, or collaborate on something,
%% easily being able to share that information. That's an obvious advantage and
%% nice to have.''

%% Kenyon: ``I like being able to look at other people's notebooks.''

\subsection{Forking}
The ability to start work where someone else let off, by forking, was popular.
Almost twice as many (131) users have forked
someone else's notebook as have starred one (75). Lilo says, ``It's one of the
handiest functions, because instead of having to find it, copy, paste it, you
just hit Fork, rename it, and it's done. It's pretty amazing.''

@@

%% ``a lot of the time we just
%% insert print statements here and there and check values. Having debugging code
%% inside a functioning program is helpful for others to understand what
%% you're doing.''

%% Leith: ``gives the audience a way to try their own analysis, like maybe Leith
%% should have used this parameter and they can just change the parameters''

A common but ill-advised use of forking is to change
parameters. Wendy says, ``I've been forking other people's notebooks [because] I
want to run them on a different part of data, or I want to change some parts,
I don't want to see this column, I want a different column, things like
that.'' Eric complains that a notebook might say ```This is a report of
the volume of all of our feeds for this month', and someone would want to
look at it for the next month or the previous month, so they'd fork it to
change the month.''

Parameters can be added to a notebook URL, but adding user interface
elements to do the same thing takes more expertise. Tool builders see a
need for facilities to make that easier. Allison calls it ``web-enabling''
notebooks: ``[Users] can always fork the notebooks and make changes, but I
feel that if the owner of the notebook web-enables it, it's easier to
run. Instead of forking it, if they can set options, it's probably more
efficient, and they can [still] fork it if they want to.''

%% Lilo: ``For the things that I'll do multiple times, I'll try to parameterize
%% it, and I'm working now on putting some packages together that will operate in
%% some fashion outside of RCloud, straight from the command line''

\subsection{Automatic source control}
%% Kenyon: ``when I've finished something, there's a nice clean record of it.''
%% (I think this may be more about the publish/markdown than about source control.)

Automatic source control proved helpful. Lilo says, ``Instead of
looking back and saying I've got a billion files here in this subdirectory and I
hope I've got them backed up, if they're on RCloud I know they are.''

%%  [...] I
%% don't have to worry about it, if I drop my computer in the bathtub, everythings
%% gone, because I'm not great about uploading my own for versioning.''

Iris points out that automatic versioning works well for coping with
the details of web development:
\begin{quote}
I like the fact that it has a built-in editor, so
if you need to fix a typo in a link, or an extra line break or other
nonsense, you can just switch to the edit view, pull up your asset, type
something, it's automatically saved, committed, everything. You don't have to go
back to your source code, change it, commit it to the repo, pull the repo to
your distribution version. 
%% None of that's necessary. A lot of web development is
%% fixing stupid little bugs, and you can do that instantly without a lot of
%% overhead, which is nice.
\end{quote}

On the other hand, saving every change leads to many fine-grained
versions. Kenyon says that for this reason, the history feature is not that
helpful: ``I don't need something that keeps track of every mistake I've made or
every direction I've tried.''
%% Tagging these versions with names wouldn't
%% necessarily help because ``namespaces are already crowded, so trying to remember
%% the names of notebooks is hard, much less the names of states within each
%% notebook.''

\subsection{Discovering others' work}
Many users find the search function valuable. Wendy says, ``The fact that we have
all the notebooks there, searchable, saves me from replicating what other people
have done.''
But other users prefer to browse just the notebooks of experts they
know. Kenyon says, ``Usually I know somebody's notebooks that I want to search
through, because I know that the kind of thing I'm looking for is something
more obscure than I'm likely to find in some random person's.''
More selective ways to search will be needed as the number of notebooks
grows further. We mention possible approaches in Section~\ref{sec:discussion}.

@@

%% Allison: ``another thing I really like about RCloud is your publish feature, so
%% that non-RCloud users, the anonymous login, non-RCloud users can access the
%% notebooks without having to login to RCloud''

%% accidental feature: ``some of my jobs are huge and I like to just call them and
%% the fact that the user can just shut down his browser or shut down his computer
%% and the job still runs in the background is actually a very neat thing. So what,
%% um, the way I have coded is, whenever I have a huge job to run, I just display
%% to the user, you go do whatever you want, we'll send you an email when the job
%% is done. That's another feature of RCloud that's very helpful, that the job
%% will continue to run in the background''

\subsection{Exploring elsewhere}

Although most analysts appreciated RCloud's sharing features,
it was not a popular tool for exploratory data analysis.
All analysts with R experience prefered to do analysis in another
tool, then paste code into RCloud for sharing.
It appears this is mainly due to familiarity.
Eric asks, ``Why would I want to use
RCloud over my current setup? If it's just me, I like my text editor and
terminal. There's nothing that I want that those two don't give me.''

%% Much of this is because of familiarity with other tools. Hugh: ``for other
%% people who use RCloud vs RStudio, they will always think that they want to
%% prototype on RStudio. Maybe because it's more familiar for them. So, they work
%% on RStudio, on a small data set, and they fix up everything, make sure
%% everything is right, and then they place it on RCloud for sharing.''

%% Many of the analysts we spoke to preferred using a text editor to refine and
%% store their work, pasting it to an R command line to try it out.

%% But there are interface reasons as well. Kenyon describes a common workflow:
%% \begin{quote}
%% I have the discipline of having a file that describes what I'm doing, with the
%% commands that I'm using, so that I can go back and recreate it or pass it on to
%% someone else. So it's a little bit like having an RCloud notebook. It's not
%% necessarily executable but most of the commands that I've typed are in there.
%% \end{quote}

@@

%% Running anything on the Web can also be a source of frustration. Poor network
%% connections can cause web sockets to die, and Kaylee complains ``What frustrates me is
%% that if I do leave this window for what feels like 5 or 7 minutes, it
%% disconnects and I have to reconnect and run the whole notebook again.'' Wendy
%% whines that ``being web-based, it's obviously slower than something you would be
%% running on your machine.'' For Joy, even the little lag of sending a command
%% to a server and receiving the result is intolerable

@@

\subsection{Cells versus the command line}
RCloud's notebook interface combines editable Markdown with a command line
interface.
The interface takes some learning. Lilo says,
\begin{quote}
%% It took me a couple of weeks of
%% looking at it to become comfortable because 
I saw the cells and it just threw me
for a total loop. I mean, it's a good idea because [...] I can run it all at
once if I want or I can break these into sections for either debugging or
staging purposes. I really like it now, but when I first saw it, [it was] very
confusing.
\end{quote}

The difference between a cell and a command prompt confuses some people.
Kenyon explains:
\begin{quote}
If I was typing into one of the cells near that top, I had to think ``I'm editing
this cell and then I have to execute it,'' and if I was editing one at the bottom,
I could type it but then it would automatically execute [by pressing enter].
%, then it became a standard cell and I have to edit it.
\end{quote}

%% For some users, the cells getting saved feels incompatible with exploratory data
%% analysis. Kenyon notes that Markdown is a form of literate programming, and
%% ``being literate about anything typically takes a lot of rewriting and going
%% back over things. [...] I just want to explore a few things and then I'll know
%% what I want to write.''

Much of the time, commands typed at the R command line serve only a transitory
purpose, so having RCloud persist commands in cells can be annoying.
Coby notes, ``It saves everything I do like everything is gold, but most
of it is junk not meant to be saved.'' Kenyon says, ``I just want to type in a
couple of quick commands and get some results that are going to tell
me what to do next, and they're not necessarily archival in any sense.''

@@

Users felt that cells do not capture the right level of granularity: they
hold either too much code or too little. Kenyon says that RCloud's cell structure
``tempts me to type a big long thing and then run the whole thing, as opposed to
typing a few little pieces and then put them together'' as he would on the command line.
When Eric uses the R command line, he ``copies and pastes 5-10 lines of code,
so when something breaks, I get an error message on that one line, and I can
up-arrow and change it and fix it, whereas in RCloud I have to run a whole cell,
so the only way to get that functionality is if every line's in one cell.''
Another complaint was that cells can take up too much vertical space, requiring scrolling.
% Rick complains, ``Every time I type more stuff, the notebook gets longer and
% longer and it's harder to deal with.'' Ivan says that cells' controls and blank
% space take extra screen space compared to a straight command line, and he and
% KC both complained that when there are long results or plots, it causes the
% code to scroll off the screen. Raif explains, ``Imagine I run a cell whose
% output is huge. [...] To find the next cell, I have to scroll down and find
% where that cell starts. So I'm losing the continuity of my code.''

Some users would like a way to keep results separate from code.
Gerrard thinks RStudio's layout is more helpful because charts
are shown in another pane that stays in place while doing analysis. Wendy says
``Cells are really useful, [but] you want to see your output in a different
window or on a separate part of the window''.

The cell structure also can be problematic if some cells take a long time to
execute. High variance is common: cells may take anywhere from seconds to
hours to run. In this situation, the Run Whole Notebook button is dangerous.
Coby reports that when converting his work to notebooks, he
ends up with a lot of comments saying ``This cell takes a long time to run.''
To avoid this pitfall, some users write code to explicitly cache results.
% Coby ends up ``littering the notebook with little switches that comment out the
% slow parts'' and either load or save the object to disk. One of the authors of this study manually writes cells that check if a result file exists, and perform a calculation only when needed.

%% Kenyon on auditing (possible solution): ``you didn't see your mistakes, they were
%% in there but you never really saw them because a mistake never led to an answer
%% at the end'' ``the real result of this analysis was these three plots, so go
%% back and figure out everything that I did that was involved in creating those
%% three plots. So that I could start from the raw data and create those three
%% plots.''

%% RCloud combines the functions of scratchpad with notebook assets. Tina thinks
%% that having a scratch pad that doesn't persist between different notebooks is
%% ``useless''.

%% The one-window design also makes some users feel that the side panes are taking
%% up too much space. Gerrard: having ``long term'' code (which would be in assets)
%% above the command prompt, like in RStudio, means that both can be sized very
%% wide to accomodate long lines.

%% Coby: Problem with screen real estate: side panes take up too much
%% space. Doesn't need the notebook tree pane.

%% Kaylee: In the previous version, it switched between code and results. Now you
%% see both - ``and there are times when that's good, but there's certainly times
%% when it'd be nice to - you know, you have 10 or 15 cells and you just want to
%% get back and there's all this output and you're scrolling all over the place so,
%% if there were a way to control both, that'd be great. There are times when I do
%% want to see the output, but there are times when I've seen enough of the same
%% thing.''

%% Allison: ``It's good that I can see the results but if I'm working and I need
%% more room and I want to hide the results, I can't.'' Could we have ``an option
%% to close the results window?''

@@

Although RCloud promises an environment where notebooks should keep working,
not all our users have learned the habits that make this a reality. As Joy
puts it, there is still a big problem with ``bitrot'' -- notebooks often
stop working because the user changed the structure of their data, or
changed a filename or a database. She says we need ``organizational
protocols'' to catch up with the technology. Even if one forks someone
else's notebook and corrects it, the original notebook still exists with
the error.

%% At first, he had to learn how to write safe
%% notebooks: ``It would be a file inside a directory that only I had [...] access
%% to and so someone else couldn't run it, [...] a temporary file on something that
%% was read-only to everyone else. But I've gotten careful about that.''

@@

The problem of dead notebooks is compounded by changing
packages and example notebooks at the same time. Eric will ``go back and
change [a notebook] and update the page, but in the meantime people
have forked it. I'll have to change a package as well, so their
old fork stops working, and they complain.''
% we don't need a new paragraph here - -scn
Eric tried to keep his notebook tree well organized, but this didn't help,
because some people kept old links in email, or forked notebooks which 
had become obsolete.
% By design, notebooks that are deleted in the user interface
% are not purged from the repository.
He says he became reluctant to share notebooks: ``Do I want to support this forever?''

>>>>

\section{Evaluation\label{sec:interviews}}

To evaluate the effectiveness of RCloud, we interviewed 13 current or
recent users of RCloud. 9 are data analysts, and 4 build tools for data
analysts and business needs. Users felt that sharing works well, but the notebook
interface slows down exploration, and the browsing interface is not scaling well.
%gw (trying to be specific)

@@

@@

Eric develops packages for analysts, and uses RCloud ``for sharing code with
other people, and for doing tutorials for {\em iotools} or {\em hmr}'', his packages.
``I want people to see how the package works, so I clearly want them
to see the code... I write it just like I would write GitHub
Markdown, where you have little code snippets and text, but RCloud lets me
actually run the snippets [and display the results].'' He also uses RCloud for
describing and debugging data sources.

Some users, who tried RCloud and were not able to continue for organizational
reasons, miss certain capabilities. Leith ``[likes] \comment{the concept of} being
able to create notebooks and share them... A wiki is not the best way for
communicating results - it's \comment{kind of} like writing a blog post with very limited
functionality. I have to save every picture and post it as an image.'' He adds,
``I can't share all of the code because it would just get crowded and wouldn't
look right on a wiki.'' Iris explains, ``If I could make a folder on RCloud and
have Python notebooks and also Pig notebooks there, and execute them from RCloud,
that would be much better than my current [environment], because that would free
me from manual documentation and version control and also telling people where
my code was. It would be just, hey, go look on RCloud, here's the stuff.''

\subsection{Forking}
The ability to fork someone's notebook and continue their work was popular, but it
should probably not be the only way to change notebooks.
Almost twice as many (131) users have forked
someone else's notebook as have starred one (75). Lilo says, ``It's one of the
handiest functions, because instead of having to find it, copy, paste it, you
just hit Fork, rename it, and it's done. It's pretty amazing.''

@@

A common but ill-advised use of forking is to change
parameters. Eric complains that a notebook might say ```This is a report of
the volume of all of our feeds for this month', and someone would want to
look at it for the next month or the previous month, so they'd fork it to
change the month.''

Parameters can be added to a notebook URL, but adding user interface
elements to do the same thing takes more expertise. Tool builders told us it
should be easier. Allison explains: ``[Users] can always fork the notebooks
and make changes, but I
feel that if the owner of the notebook [adds a user interface], it's easier to
run. Instead of forking it, if they can set options, it's probably more
efficient, and they can [still] fork it if they want to.''

\subsection{Automatic source control}
Automatic source control is helpful, but it also leads to many commits. Lilo says, ``Instead of
looking back and saying I've got a billion files here in this subdirectory and I
hope I've got them backed up, if they're on RCloud I know they are.''

Iris points out that automatic versioning works well for coping with
the details of web development:
\begin{quote}
I like the fact that it has a built-in editor, so
if you need to fix a typo in a link, or an extra line break or other
nonsense, you can just switch to the edit view, pull up your asset, type
something, it's automatically saved, committed, everything. You don't have to go
back to your source code, change it, commit it to the repo, pull the repo to
your distribution version.
\end{quote}

On the other hand, saving every change leads to many fine-grained
versions. Kenyon says that for this reason, the history feature is not that
helpful: ``I don't need something that keeps track of every mistake I've made or
every direction I've tried.''

\subsection{Discovering others' work}
Users reported that the search function is more helpful for learning about
particular functions than techniques. Wendy says, ``The fact that we have
all the notebooks there, searchable, saves me from replicating what other people
have done.''
But other users prefer to browse just the notebooks of experts they
know. Kenyon says, ``Usually I know somebody's notebooks that I want to search
through, because \comment{I know that} the kind of thing I'm looking for is something
more obscure than I'm likely to find in some random person's.''
More selective ways to search will be needed as the number of notebooks
grows further. We mention possible approaches in Section~\ref{sec:discussion}.

@@

\subsection{Blocked by the web}

Although most analysts appreciated RCloud's sharing features,
it was not a popular tool for exploratory data analysis.
All analysts with R experience prefered to do analysis in another
tool, then paste code into RCloud for sharing, because the web interface got in the
way of their work.

@@

@@

\subsection{Cells versus the command line}
RCloud's notebook interface combines editable Markdown with a command line
interface. Many users felt that this format, so effective for presentation,
is incompatible with exploratory data analysis.

Much of the time, commands typed at the R command line serve only a transitory
purpose, so having RCloud persist commands in cells can be annoying.
Coby notes, ``It saves everything I do like everything is gold, but most
of it is junk not meant to be saved.''

@@

Users felt that cells do not capture the right level of granularity.
Kenyon says that RCloud's cell structure
``tempts me to type a big long thing and then run the whole thing, as opposed to
typing a few little pieces and then put them together'' as he would on the command line.
When Eric uses the R command line, he ``copies and pastes 5-10 lines of code,
so when something breaks, I get an error message on that one line, and I can
up-arrow and change it and fix it, whereas in RCloud I have to run a whole cell,
so the only way to get that functionality is if every line's in one cell.''

\subsection{Too vertical}
The strict vertical linearity of a notebook also poses space and time problems
not encountered on the command line, where typically only the last few lines are looked.

Scrolling between code and results is time consuming, and some users would
prefer to keep results separate from code.
Gerrard thinks RStudio's layout is more helpful because charts
are shown in another pane that stays in place while doing analysis. Wendy says
``Cells are really useful, [but] you want to see your output in a different
window or on a separate part of the window''.

The cell structure also can be problematic if some cells take a long time to
execute. High variance is common: cells may take anywhere from seconds to
hours to run. In this situation, the Run Whole Notebook button is dangerous.
Coby reports that when converting his work to notebooks, he
ends up with a lot of comments saying ``This cell takes a long time to run.''
To avoid this pitfall, some users write code to explicitly cache results.

@@

Although RCloud promises an environment where notebooks should keep working,
not all our users have learned the habits that make this a reality. As Joy
puts it, ``bitrot'' is still a big problem -- notebooks often
stop working because the user changed the structure of their data, or
changed a filename or a database. Even if one forks someone
else's notebook and corrects it, the original notebook still exists with
the error. She says we need ``organizational protocols'' to catch up with the technology.

@@

Eric tried to keep his notebook
tree well organized, but this didn't help, because people kept old links
in email, or forked notebooks which had become obsolete.
He says he became scared to share notebooks: ``Do I want to support this forever?''
